Bayesian multi-proxy reconstruction of latitudinal temperature gradients: An early Eocene example

### 2023-04-23 manuscript/main.qmd
work on intro and update fig. 2

Reconstructing past climate is fundamental to understanding the long-term evolution of the Earth's climate system, including natural climate variability [@zachos2008], the impacts of external forcings such as impact events [@chiarenza2020], and changes in greenhouse gas concentrations [@foster2017]. Moreover, palaeoclimatic reconstructions are vital for contextualising current global warming, enabling evaluation of the magnitude and significance in a broader historical perspective [@royer2004; @Burke2018], as well as understanding ecosystem response to environmental change [@jones2022reef; @woodhouse2023].

To reconstruct palaeoclimatic conditions, researchers use a range of approaches including geochemical proxies, geological evidence, and Earth System Models [see @Burgener2023]. Geochemical proxies such as oxygen isotopes ($\delta^18^O$) are used to provide quantitative palaeotemperature estimates by applying transfer functions derived from experimental work and field observations [e.g. @Liu2009; @Wuchter2004]. Geological evidence such as lithological and fossil data are used to qualitatively infer past climatic conditions based on analogous modern sediments [@Chandra2021], or the premise that the climatic requirements of ancient taxa, biological traits, or ecological communities were similar to those of their nearest modern relatives [@Peppe2011; @Royer2012; @Salonen2019]. Finally, Earth System Models (ESMs)--such as the HadCM3 climate model [@valdes2017]--are used to model the main processes of the Earth system to provide spatially and temporally explicit quantitative estimates of palaeoclimatic conditions. Nevertheless, ESMs ultimately rely on constraints provided by the geological and fossil record [@Lunt2021].

Recent advances in deep-time climate reconstructions [e.g. @Burgener2023; @chevalier2022], including improvements in proxy data quantification [e.g. @bao2023] and data synthesis [e.g. @Veizer2015; @Hollis2019; @Scotese2021; @Grossman2022; @Judd2022], are enabling researchers to better understand the complex and dynamic nature of the Earth's climate system over geologic time. However, a fundamental challenge remains which must be addressed to exploit the full potential of these advances: the proxy record of past climate is spatially biased, impacted by incomplete preservation and uneven sampling [@judd2020; @Jones2022; @Judd2022]. Recent work has demonstrated that this bias can lead to spurious estimates of latitudinal temperature gradients and global temperatures [@judd2020; @Jones2022].

Inferring regional or global-scale patterns from an incomplete record relies on the bridging of spatial gaps. A common approach that requires no additional computation is the spatial visualisation of proxy-derived temperatures against latitude, showing broad latitudinal temperature trends [@Hollis2019; @Vickers2021]. Interpolation is sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994]. Adding to this, some proxy-based reconstructions use statistical modelling to infer climatic patterns. For example, polynomial regression [@Bijl2009] and cosine functions [@Inglis2020] have been used to reconstruct latitudinal temperature gradients, and 2D-reconstructions of surface temperatures have been created with Gaussian process regression [@Inglis2020]. These approaches work well for interpolating relatively well-sampled data, but the absence of constraints on the modeled parameters means that such models can produce unrealistic temperature estimates when extrapolating from sparsely sampled data. Statistical modelling in the Bayesian framework helps overcome this problem by requiring the explicit specification of priors for the model parameters, which can be used to express physical constraints [@Chandra2021].

Here, we present a novel Bayesian hierarchical model that combines quantitative proxies and ecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data. This model expands upon existing, spatially explicit reconstructions of palaeoclimate by allowing for the integration of 1) prior information based on physical principles and on the observed, modern sea surface temperature distribution, and of 2) geochemical and ecological climate proxies in a common, quantitative framework. We use a generalised logistic function to infer the shape of the temperature gradient despite a patchy latitudinal coverage, and test the robustness of this method using a down-sampled record of modern SSTs. We apply this model to the record of the early Eocene climatic optimum (EECO), combining a compilation of geochemical proxies [@Hollis2019], mangrove communities [@Popescu2021], and coral reefs [@Zamagni2012], using a nearest-living-relative approach [see e.g. @Greenwood2017] to establish broad temperature ranges for the ecological data.

@@

Model reconstructions of simulated latitudinal temperature gradients at various sample sizes. Each column depicts a different reconstruction for given sample sizes: 5, 10, 20, and 34 (sample size for the EECO). Each row depicts a different simulated latitudinal temperature gradient that represents idealised climatic states: extreme greenhouse, greenhouse, icehouse, extreme icehouse. The black line illustrates the simulated gradient. The blue line depicts the reconstructed gradient represented by the median sea surface temperature value estimated from 1000 model runs. The blue shading depicts the 90%, 95%, and 99% credible intervals. Each gradient is depicted in absolute latitude.
:::

>>>>

Reconstructing past climate is essential to understanding the long-term evolution of the Earth's climate system and contextualising current global warming [@royer2004; @Burke2018]. Geochemical proxies, such as oxygen isotopes ($\delta^18^O$), are regularly used to provide quantitative palaeotemperature estimates for the geological past at a range of temporal and spatial scales [e.g. @Liu2009; @Wuchter2004]. Recent advances in the synthesis of such data [e.g. @Veizer2015; @Hollis2019; @Scotese2021; @Grossman2022; @Judd2022] are enabling a better understanding of the complex and dynamic nature of the Earth's climate system. However, a fundamental challenge remains which must be addressed to exploit the full potential of these advances; the proxy record of past climate is spatially incomplete and biased by imperfect preservation and uneven sampling [@judd2020; @Jones2022; @Judd2022].

Whilst geochemical proxy data can provide robust estimates of palaeotemperature at local scales (e.g. REF), recent work has demonstrated that spatial biases in the geochemical proxy record can lead to spurious estimates of regional (e.g. latitudinal temperature gradients) and global temperatures [@judd2020; @Jones2022]. At the most fundamental level, this can be driven by two factors: (1) missing data for some regions (e.g. no high-latitude data); or (2) overrepresentation of some regions (e.g. a high proportion of samples from tropical regions). The latter can be addressed through the down-sampling of data or restricting analyses to specifics regions (e.g. REFs). However, in order to robustly infer regional or global-scale patterns from an incomplete record, spatial gaps must ultimately be bridged. One common approach, that requires no additional computation, is the spatial visualisation of proxy-derived temperatures against latitude, showing broad latitudinal temperature trends [@Hollis2019; @Vickers2021]. Interpolation is also sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994]. Adding to this, some proxy-based reconstructions use statistical modelling to infer climatic patterns. For example, polynomial regression [@Bijl2009] and cosine functions [@Inglis2020] have been used to reconstruct latitudinal temperature gradients, and 2D-reconstructions of surface temperatures have been created with Gaussian process regression [@Inglis2020]. These approaches work well for interpolating relatively well-sampled data, but the absence of constraints on the modelled parameters means that such models can produce unrealistic temperature estimates when extrapolating from sparsely sampled data. Statistical modelling in a Bayesian framework can help overcome this problem by requiring the explicit specification of priors for the model parameters, which can be used to express physical constraints [@Chandra2021].

Spatial gaps in the record of past climate can also be addressed through the integration of additional data. For example, lithological and fossil data can be used to infer past climatic conditions based on analogous modern sediments [@Chandra2021], or the premise that the climatic requirements of ancient taxa, biological traits, or ecological communities were similar to those of their nearest modern relatives [@Peppe2011; @Royer2012; @Salonen2019]. Despite this potential, the integration of geochemical proxy data with other sources of information (e.g. ecological data) is rare due to the absence of a quantitative framework [@Burgener2023].

Here, we present a novel Bayesian hierarchical model that combines quantitative proxies and ecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled climate data. This model expands upon existing, spatially explicit reconstructions of palaeoclimate by allowing for the integration of (1) prior information based on physical principles and on the observed, modern sea surface temperature distribution, and of (2) geochemical and ecological climate proxies in a common, quantitative framework. We use a generalised logistic function to infer the shape of the temperature gradient despite a patchy latitudinal coverage, and test the robustness of this method using a down-sampled record of modern SSTs. We apply this model to the record of the early Eocene climatic optimum (EECO), combining a compilation of geochemical proxies [@Hollis2019], mangrove communities [@Popescu2021], and coral reefs [@Zamagni2012], using a nearest-living-relative approach [see e.g. @Greenwood2017] to establish broad temperature ranges for the ecological data. Whilst we primarily do so to demonstrate an application of the model, the EECO represents an interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], and a potential analogue for extreme climate warming scenarios [@Burke2018].

@@

Model reconstructions of simulated latitudinal temperature gradients at various sample sizes. Each column depicts a different reconstruction for given sample sizes: 5, 10, 20, and 34 (sample size for the EECO). Each row depicts a different simulated latitudinal temperature gradient that represents idealised climatic states: extreme greenhouse, greenhouse, icehouse, extreme icehouse. The black line illustrates the simulated gradient. The blue line depicts the reconstructed gradient represented by the median sea surface temperature value estimated from 1000 model runs. The blue shading depicts the 90%, 95%, and 99% credible intervals. Bold black text within each panel depicts the coefficient of determination (*R*^2^ value) for estimating goodness of fit between the simulated and modelled gradient. The median (50%) *R*^2^ value along with the 2.5% and 97.5% confidence intervals from the 1000 model runs are shown. Each gradient is depicted in absolute latitude.
:::
